---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MPCurver

<!-- badges: start -->
<!-- badges: end -->

`MPCurver` fits smooth latent orderings with Gaussian-mixture / GMRF models.

The public workflow is **`mpcurve`-first and CAVI-first**:
use `fit_mpcurve()` to fit a model, `do_mpcurve()` to continue a fitted model,
and the usual `print()`, `summary()`, and `plot()` methods to inspect results.

## Documentation

The package documentation is organized around the public `mpcurve` workflow.

- `vignettes/mpcurve_intro.Rmd`: current single-ordering introduction.
- `vignettes/partition.Rmd`: current multi-ordering / feature-partition tools.
- `vignettes/fitness.rmd`: applied case study using the public interface.

## Installation

You can install the development version of `MPCurver` like so:

``` r
pak::pak("AgueroZZ/MPCurver")
```

## Example

This is a basic example using the recommended `mpcurve` interface:

```{r setup}
library(MPCurver)
```

We will simulate a moderately easy 2D spiral and fit the default MPCurver
workflow.

```{r}
sim <- simulate_spiral2d(n = 1500, turns = 1, noise = 0.08, seed = 123)
plot(sim$obs, pch = 16, cex = 0.35, col = "grey80",
     xlab = "x1", ylab = "x2", main = "Noisy 2D spiral")
```

Now we fit the model with `fit_mpcurve()`. The public wrapper uses the CAVI
backend and returns an `mpcurve` object.

```{r}
set.seed(123)
fit <- fit_mpcurve(
  X = as.matrix(sim$obs),
  method = "isomap",
  K = 40,
  rw_q = 2,
  iter = 150,
  discretization = "equal",
  tol = 1e-8
)

plot(fit)
plot(fit, plot_type = "elbo")
```

We can compare several initialization strategies under the same `cavi` model:

```{r}
fits <- fit_mpcurve(
  X = as.matrix(sim$obs),
  method = c("PCA", "fiedler", "isomap"),
  K = 40,
  rw_q = 2,
  iter = 150,
  discretization = "equal",
  tol = 1e-8,
  num_cores = 1
)

attr(fits, "summary")
```

`fit_mpcurve()` returns either a single `mpcurve` fit or a list of fits. Each
fit keeps the unified `params`, `gamma`, `elbo_trace`, `loglik_trace`, and the
underlying backend fit in `$fit`.

For multi-ordering analyses, use `fit_mpcurve(..., intrinsic_dim = 2)` or more
generally `fit_mpcurve(..., intrinsic_dim = M)`. The same `mpcurve` methods
continue to work for both single-ordering and partition fits.
