---
title: "Analysis on the eLife Fitness Dataset using MPCurver"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

This note revisits the eLife fitness dataset using the current unified interface in `MPCurver`, with a specific focus on **learning the
number of latent ordering systems**.

We now keep the workflow intentionally simple:

- use `isomap` as the initialization method;
- set `K = 50`;

## First look at the data:

```{r load, include=FALSE, message=FALSE}
library(MPCurver)

locate_elife_csv <- function() {
    pkg_path <- system.file("extdata", "elife-61271-fig2-data1-v2.csv", package = "MPCurver")
    if (nzchar(pkg_path) && file.exists(pkg_path)) {
        return(pkg_path)
    }
    src_path <- file.path("..", "inst", "extdata", "elife-61271-fig2-data1-v2.csv")
    return(src_path)
}

csv_path <- locate_elife_csv()
dat <- read.csv(csv_path, check.names = FALSE)

fitness_cols <- grep("_fitness$", names(dat), value = TRUE)
error_cols <- grep("_error$", names(dat), value = TRUE)

X <- as.matrix(dat[, fitness_cols, drop = FALSE])
mode(X) <- "numeric"
colnames(X) <- sub("_fitness$", "", fitness_cols)

meta <- dat[, c("barcode", "gene", "type", "ploidy", "class", "mutation_type"), drop = FALSE]
```

```{r helpers, include=FALSE}
environment_category <- function(env) {
    if (grepl("Ferm|Day", env)) {
        return("time course")
    }
    if (grepl("NaCl|KCl", env)) {
        return("salt stress")
    }
    if (grepl("Ben|Flu|GdA|DMSO", env)) {
        return("drug / chemical")
    }
    if (grepl("Gluc|Baffle", env)) {
        return("glucose / batch")
    }
    if (grepl("Raf|Gly|EtOH|Suc", env)) {
        return("carbon source")
    }
    "other"
}

env_category <- vapply(colnames(X), environment_category, character(1))

plot_ordered_heatmap <- function(X, order_idx, env_idx = seq_len(ncol(X)),
                                 main = "", zlim = c(-2.5, 2.5)) {
    Xsub <- X[order_idx, env_idx, drop = FALSE]
    Xsub <- scale(Xsub)
    Xsub[!is.finite(Xsub)] <- 0
    pal <- colorRampPalette(c("#313695", "#74add1", "#f7f7f7", "#f46d43", "#a50026"))(101)
    image(
        x = seq_len(nrow(Xsub) + 1L),
        y = seq_len(ncol(Xsub) + 1L),
        z = Xsub,
        col = pal,
        axes = FALSE,
        xlab = "Mutants ordered by inferred pseudotime",
        ylab = "",
        zlim = zlim,
        main = main
    )
    axis(2, at = seq_len(ncol(Xsub)), labels = colnames(Xsub), las = 2, cex.axis = 0.55)
    box()
}

plot_pi_profile <- function(pi_vec, main) {
    barplot(
        pi_vec,
        border = NA,
        col = "#4477AA",
        main = main,
        xlab = "Latent position",
        ylab = expression(hat(pi)[k])
    )
}

plot_representative_trajectories <- function(X, fit_partition, which_assign = "A",
                                             n_show = 6L, main_prefix = "") {
    ord_idx <- if (which_assign == "A") 1L else 2L
    fit_ord <- fit_partition$fits[[ord_idx]]$fit
    pt_map <- fit_partition$locations[[which_assign]]$map$index
    pt_mean <- fit_partition$locations[[which_assign]]$mean$pseudotime
    assign_vec <- fit_partition$partition$assign

    env_idx <- which(assign_vec == which_assign)
    if (!length(env_idx)) {
        return(invisible(NULL))
    }

    cor_vec <- vapply(env_idx, function(j) {
        suppressWarnings(cor(X[, j], pt_mean, method = "spearman"))
    }, numeric(1))
    env_idx <- env_idx[order(abs(cor_vec), decreasing = TRUE)]
    env_idx <- env_idx[seq_len(min(n_show, length(env_idx)))]

    mu <- fit_ord$posterior$mean[env_idx, , drop = FALSE]
    K <- ncol(mu)
    obs_mean <- t(vapply(env_idx, function(j) {
        out <- rep(NA_real_, K)
        for (k in seq_len(K)) {
            idx <- which(pt_map == k)
            if (length(idx)) out[k] <- mean(X[idx, j])
        }
        out
    }, numeric(K)))

    nr <- ceiling(length(env_idx) / 2)
    op <- par(mfrow = c(nr, 2), mar = c(3.5, 3.5, 2.5, 1))
    on.exit(par(op), add = TRUE)
    for (ii in seq_along(env_idx)) {
        jj <- env_idx[ii]
        ylim <- range(c(mu[ii, ], obs_mean[ii, ]), na.rm = TRUE)
        plot(
            seq_len(K), obs_mean[ii, ],
            type = "o",
            pch = 16,
            col = "#666666",
            ylim = ylim,
            xlab = "Latent position",
            ylab = "Fitness",
            main = paste0(main_prefix, colnames(X)[jj])
        )
        lines(seq_len(K), mu[ii, ], type = "o", pch = 19, col = "#D95F02", lwd = 2)
        legend("topleft",
            legend = c("Observed bin mean", "Posterior mean"),
            col = c("#666666", "#D95F02"),
            lty = 1, pch = c(16, 19), bty = "n", cex = 0.8
        )
    }
}

trajectory_roughness <- function(raw_fit) {
    M <- raw_fit$posterior$mean
    Q <- raw_fit$Q_K
    mean(vapply(seq_len(nrow(M)), function(j) {
        as.numeric(t(M[j, ]) %*% Q %*% M[j, ])
    }, numeric(1)))
}
```

Take a look at the number of mutants and environments.
```{r data-summary}
data.frame(
    quantity = c("mutants", "environments", "fitness columns", "error columns"),
    value = c(nrow(X), ncol(X), length(fitness_cols), length(error_cols))
) |>
    knitr::kable()

sort(table(env_category), decreasing = TRUE) |>
    as.data.frame() |>
    stats::setNames(c("category", "n_environments")) |>
    knitr::kable()
```


## Initial fit of a MPCurver

First, we will try to fit a MPCurve using `MPCurver` package, the goal is to learn the latent position of all the mutants such that their fitness scores are functions that vary smoothly along the latent positions.

```{r uni-fits}
fit_unidir <- fit_mpcurve(
    X,
    method = "isomap",
    K = 50,
    rw_q = 2,
    verbose = FALSE
)
```

Take a look at the convergence:

```{r convergence, fig.width = 8, fig.height = 8, out.width = "100%"}
plot(fit_unidir, plot_type = "elbo")
```

Look at some fitted trajectories (first 4 environments):

```{r uni-trajectory, fig.width = 8, fig.height = 8, out.width = "100%"}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(fit_unidir, dims = 1)
plot(fit_unidir, dims = 2)
plot(fit_unidir, dims = 3)
plot(fit_unidir, dims = 4)
par(mfrow = c(1, 1))
```

These four environments (`r paste(colnames(X)[1:4], collapse = ", ")`) show very similar trajectories along the latent position, both appearing to increase with latent positions. 
However, the fitted curves appear to be quite wiggly. 
To make the curves smoother, we can apply a prior on $\lambda$, which controls the roughness of the trajectories. 
The prior is an exponential prior assigned to $1/sqrt(\lambda)$. 
By increasing the rate parameter of the exponential prior, we can make the fitted trajectories smoother.

```{r uni-fits-smooth}
fit_unidir_smooth <- fit_mpcurve(
    X,
    method = "isomap",
    K = 50,
    iter = 500,
    rw_q = 2,
    lambda_sd_prior_rate = 1e4,
    verbose = FALSE
)
```

Again, check if the model has converged:

```{r uni-convergence-smooth, fig.width = 8, fig.height = 8, out.width = "100%"}
plot(fit_unidir_smooth, plot_type = "elbo")
summary(fit_unidir_smooth)
```

Take a look at the fitted trajectories:

```{r uni-trajectory-smooth, fig.width = 8, fig.height = 8, out.width = "100%"}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(fit_unidir_smooth, dims = 1)
plot(fit_unidir_smooth, dims = 2)
plot(fit_unidir_smooth, dims = 3)
plot(fit_unidir_smooth, dims = 4)
par(mfrow = c(1, 1))
```

This looks much more reaonsable to me. 
We can compare the mean latent positions learned by the two models:

```{r compare-latent-positions, fig.width = 8, fig.height = 8, out.width = "100%"}
cor(fit_unidir_smooth$locations$mean$pseudotime, fit_unidir$locations$mean$pseudotime)
plot(fit_unidir_smooth$locations$mean$pseudotime, fit_unidir$locations$mean$pseudotime,
    xlab = "Latent position (smooth)", ylab = "Latent position (rough)"
)
abline(0, 1, lty = 2)
```

The latent positions do not change much, but the trajectories are much smoother. 
