---
title: "Analyzing the eLife Fitness Dataset with MPCurver"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Analyzing the eLife Fitness Dataset with MPCurver}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5.5
)
```

## Overview

This vignette shows how to use the public `fit_mpcurve()` interface on the
eLife fitness dataset. The goal is not just to produce a fit, but to explain
the modeling logic step by step.

We will use `MPCurver` to answer three practical questions:

1. Can a single latent ordering explain mutant fitness across environments?
2. If we encourage smoother trajectories, do we keep the same ordering while
   improving interpretability?
3. Do the data support multiple ordering systems, or does a single ordering
   remain sufficient once we account for measurement error?

The recommended workflow throughout is the default `cavi` backend exposed by
`fit_mpcurve()`.

## Load package and data

```{r load}
if (requireNamespace("MPCurver", quietly = TRUE)) {
  library(MPCurver)
} else if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(".", quiet = TRUE)
} else {
  stop("Need either an installed MPCurver package or devtools::load_all().")
}

set.seed(42)

locate_elife_csv <- function() {
  pkg_path <- system.file("extdata", "elife-61271-fig2-data1-v2.csv", package = "MPCurver")
  if (nzchar(pkg_path) && file.exists(pkg_path)) {
    return(pkg_path)
  }

  src_path <- file.path("inst", "extdata", "elife-61271-fig2-data1-v2.csv")
  if (file.exists(src_path)) {
    return(src_path)
  }

  stop("Could not locate elife-61271-fig2-data1-v2.csv.")
}

csv_path <- locate_elife_csv()
dat <- read.csv(csv_path, check.names = FALSE)

fitness_cols <- grep("_fitness$", names(dat), value = TRUE)
error_cols <- grep("_error$", names(dat), value = TRUE)

X <- as.matrix(dat[, fitness_cols, drop = FALSE])
mode(X) <- "numeric"
colnames(X) <- sub("_fitness$", "", fitness_cols)

S <- as.matrix(dat[, error_cols, drop = FALSE])
mode(S) <- "numeric"
colnames(S) <- sub("_error$", "", error_cols)

meta <- dat[, c("barcode", "gene", "type", "ploidy", "class", "mutation_type"), drop = FALSE]
```


Just using the error columns as the true SEs might be too risky, so we will adjust the SEs using t-statistics-adjustment:

```{r adjust-se}
t_adjust_se <- function(Y, S, nu, tol = 1e-12) {
  if (!is.matrix(Y) || !is.numeric(Y)) {
    stop("`Y` must be a numeric matrix.")
  }
  if (!is.matrix(S) || !is.numeric(S)) {
    stop("`S` must be a numeric matrix.")
  }
  if (!all(dim(Y) == dim(S))) {
    stop("`Y` and `S` must have the same dimensions.")
  }
  if (any(S < 0, na.rm = TRUE)) {
    stop("`S` must be nonnegative.")
  }
  
  if (length(nu) == 1L) {
    if (!is.numeric(nu) || is.na(nu) || nu <= 0) {
      stop("`nu` must be a positive number.")
    }
    nu_mat <- matrix(nu, nrow = nrow(Y), ncol = ncol(Y))
  } else {
    if (!is.matrix(nu) || !is.numeric(nu) || !all(dim(nu) == dim(Y))) {
      stop("`nu` must be either a positive scalar or a numeric matrix with the same dimensions as `Y`.")
    }
    if (any(nu <= 0, na.rm = TRUE)) {
      stop("All entries of `nu` must be positive.")
    }
    nu_mat <- nu
  }
  
  z_t <- Y / S
  
  # Main calculation:
  # S_tilde = S * z_t / qnorm(pt(z_t, df = nu))
  p_t <- pt(z_t, df = nu_mat)
  z_n <- qnorm(p_t)
  
  S_tilde <- S * z_t / z_n
  
  # Numerical fix for z_t close to 0:
  # limit of z / qnorm(pt(z, df=nu)) as z -> 0 is dnorm(0) / dt(0, df=nu)
  near_zero <- is.finite(z_t) & abs(z_t) < tol
  if (any(near_zero, na.rm = TRUE)) {
    limit_factor <- dnorm(0) / dt(0, df = nu_mat[near_zero])
    S_tilde[near_zero] <- S[near_zero] * limit_factor
  }
  
  # Preserve zeros
  zero_se <- (S == 0) & !is.na(S)
  if (any(zero_se)) {
    S_tilde[zero_se] <- 0
  }
  
  dimnames(S_tilde) <- dimnames(S)
  S_tilde
}
S <- t_adjust_se(X, S, nu = 4)
```



In this dataset:

- each row is a mutant;
- each column is an environment;
- each matrix entry is a fitness score;
- and the paired `*_error` columns provide observation-level measurement
  standard deviations that we can optionally pass to `fit_mpcurve()` via `S`.

## First look at the data

```{r helpers, include=FALSE}
environment_category <- function(env) {
  if (grepl("Ferm|Day", env)) return("time course")
  if (grepl("NaCl|KCl", env)) return("salt stress")
  if (grepl("Ben|Flu|GdA|DMSO", env)) return("drug / chemical")
  if (grepl("Gluc|Baffle", env)) return("glucose / batch")
  if (grepl("Raf|Gly|EtOH|Suc", env)) return("carbon source")
  "other"
}

env_category <- vapply(colnames(X), environment_category, character(1))

plot_ordered_heatmap <- function(X, order_idx, env_idx = seq_len(ncol(X)),
                                 main = "", zlim = c(-2.5, 2.5)) {
  Xsub <- X[order_idx, env_idx, drop = FALSE]
  Xsub <- scale(Xsub)
  Xsub[!is.finite(Xsub)] <- 0
  pal <- colorRampPalette(c("#313695", "#74add1", "#f7f7f7", "#f46d43", "#a50026"))(101)
  image(
    x = seq_len(nrow(Xsub) + 1L),
    y = seq_len(ncol(Xsub) + 1L),
    z = Xsub,
    col = pal,
    axes = FALSE,
    xlab = "Mutants ordered by inferred pseudotime",
    ylab = "",
    zlim = zlim,
    main = main
  )
  axis(2, at = seq_len(ncol(Xsub)), labels = colnames(Xsub), las = 2, cex.axis = 0.55)
  box()
}

plot_representative_trajectories <- function(X, fit_partition, which_assign = "A",
                                             n_show = 6L, main_prefix = "") {
  ord_idx <- match(which_assign, names(fit_partition$fits))
  if (is.na(ord_idx)) {
    return(invisible(NULL))
  }

  fit_ord <- fit_partition$fits[[ord_idx]]$fit
  pt_map <- fit_partition$locations[[which_assign]]$map$index
  pt_mean <- fit_partition$locations[[which_assign]]$mean$pseudotime
  assign_vec <- fit_partition$partition$assign

  env_idx <- which(assign_vec == which_assign)
  if (!length(env_idx)) {
    return(invisible(NULL))
  }

  cor_vec <- vapply(env_idx, function(j) {
    suppressWarnings(cor(X[, j], pt_mean, method = "spearman"))
  }, numeric(1))
  env_idx <- env_idx[order(abs(cor_vec), decreasing = TRUE)]
  env_idx <- env_idx[seq_len(min(n_show, length(env_idx)))]

  mu <- fit_ord$posterior$mean[env_idx, , drop = FALSE]
  K <- ncol(mu)
  obs_mean <- t(vapply(env_idx, function(j) {
    out <- rep(NA_real_, K)
    for (k in seq_len(K)) {
      idx <- which(pt_map == k)
      if (length(idx)) out[k] <- mean(X[idx, j])
    }
    out
  }, numeric(K)))

  nr <- ceiling(length(env_idx) / 2)
  op <- par(mfrow = c(nr, 2), mar = c(3.5, 3.5, 2.5, 1))
  on.exit(par(op), add = TRUE)
  for (ii in seq_along(env_idx)) {
    jj <- env_idx[ii]
    ylim <- range(c(mu[ii, ], obs_mean[ii, ]), na.rm = TRUE)
    plot(
      seq_len(K), obs_mean[ii, ],
      type = "o",
      pch = 16,
      col = "#666666",
      ylim = ylim,
      xlab = "Latent position",
      ylab = "Fitness",
      main = paste0(main_prefix, colnames(X)[jj])
    )
    lines(seq_len(K), mu[ii, ], type = "o", pch = 19, col = "#D95F02", lwd = 2)
    legend("topleft",
      legend = c("Observed bin mean", "Posterior mean"),
      col = c("#666666", "#D95F02"),
      lty = 1, pch = c(16, 19), bty = "n", cex = 0.8
    )
  }
}
```

```{r data-summary}
data.frame(
  quantity = c("mutants", "environments", "fitness columns", "error columns"),
  value = c(nrow(X), ncol(X), length(fitness_cols), length(error_cols))
) |>
  knitr::kable()

sort(table(env_category), decreasing = TRUE) |>
  as.data.frame() |>
  stats::setNames(c("category", "n_environments")) |>
  knitr::kable()
```

This is a good MPCurver example because the environments span several broad
condition types, while the mutants are shared across all columns. That makes it
natural to ask whether the same latent mutant ordering explains every
environment, or whether different subsets of environments prefer different
orderings.

## Step 1: Fit a single-ordering MPCurve

We start with the simplest question: can one latent ordering explain the main
structure in the dataset?

The call below uses the main user-facing API:

- `method = "isomap"` supplies the initialization;
- `K = 50` uses a 50-bin latent grid;
- `rw_q = 2` uses a second-order random-walk prior for smooth trajectories.

```{r uni-fit}
fit_unidir <- fit_mpcurve(
  X,
  method = "isomap",
  K = 50,
  rw_q = 2,
  verbose = FALSE
)
```

### Check convergence

For the default `cavi` backend, the main quantity to inspect is the ELBO trace.

```{r uni-convergence, out.width = "100%"}
plot(fit_unidir, plot_type = "elbo")
```

### Inspect a few fitted trajectories

```{r uni-trajectories, fig.height = 8, out.width = "100%"}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(fit_unidir, dims = 1)
plot(fit_unidir, dims = 2)
plot(fit_unidir, dims = 3)
plot(fit_unidir, dims = 4)
par(mfrow = c(1, 1))
```

The first fit already recovers coherent structure, but the trajectories are
still rougher than we would like for interpretation. 
This is a common pattern on real data: the latent ordering may be sensible even when the fitted curves need stronger regularization.

## Step 2: Encourage smoother trajectories

We now refit the same single-ordering model, but add a stronger prior on
`lambda_j` through `lambda_sd_prior_rate`. Larger values encourage smoother
trajectories.

```{r uni-fit-smooth}
fit_unidir_smooth <- fit_mpcurve(
  X,
  method = "isomap",
  K = 50,
  iter = 500,
  rw_q = 2,
  lambda_sd_prior_rate = 1e4,
  verbose = FALSE
)
```

```{r uni-convergence-smooth, out.width = "100%"}
plot(fit_unidir_smooth, plot_type = "elbo")
fit_unidir_smooth
```

```{r uni-trajectories-smooth, fig.height = 8, out.width = "100%"}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(fit_unidir_smooth, dims = 1)
plot(fit_unidir_smooth, dims = 3)
plot(fit_unidir_smooth, dims = 7)
plot(fit_unidir_smooth, dims = 9)
par(mfrow = c(1, 1))
```

These curves are easier to read: the global ordering remains, but the fitted
environment trajectories are less wiggly.

A useful sanity check is whether smoothing changes the inferred mutant ordering
itself, or mainly changes the trajectory estimates.

```{r compare-single-orderings, out.width = "100%"}
cor(
  fit_unidir_smooth$locations$mean$pseudotime,
  fit_unidir$locations$mean$pseudotime
)

plot(
  fit_unidir_smooth$locations$mean$pseudotime,
  fit_unidir$locations$mean$pseudotime,
  xlab = "Latent position (smoother fit)",
  ylab = "Latent position (less regularized fit)"
)
abline(0, 1, lty = 2)
```

In this dataset the two fits stay close, which is reassuring: stronger
regularization mostly improves interpretability rather than changing the basic
ordering story.

## Step 3: Ask whether one ordering is enough

The single-ordering fit assumes that every environment varies smoothly along
the same latent mutant axis. That may be too restrictive. Some environments may
reflect a different adaptation program and therefore prefer a different latent
ordering.

To test that idea, we let `fit_mpcurve()` search over up to four latent
ordering systems using greedy backward selection. 
The model starts with four orderings and prunes back to the optimal number based on the ELBO.

```{r partition-fit}
fit_partition <- fit_mpcurve(
  X,
  method = "isomap",
  intrinsic_dim = 4,
  greedy = "backward",
  ridge = 1e-6,
  K = 50,
  iter = 500,
  rw_q = 2,
  lambda_sd_prior_rate = 1e4,
  verbose = FALSE
)
```

```{r partition-summary}
fit_partition
```

### Look at some fitted orderings

```{r partition-plot, out.width = "100%"}
par(mfrow = c(2, 1), mar = c(4, 4, 3, 1))
plot(fit_partition, dims = 1)
plot(fit_partition, dims = 27)
par(mfrow = c(1, 1))
```

### Which environments were assigned to which ordering?

```{r partition-table}
assignment_tbl <- data.frame(
  environment = colnames(X),
  category = env_category,
  ordering = fit_partition$partition$assign,
  stringsAsFactors = FALSE
)

assignment_tbl[order(assignment_tbl$ordering, assignment_tbl$category, assignment_tbl$environment), ] |>
  knitr::kable(row.names = FALSE)
```

### How similar are the inferred orderings?

```{r partition-correlation}
ordering_names <- names(fit_partition$locations)
ordering_names <- ordering_names[ordering_names %in% unique(fit_partition$partition$assign)]

t_df <- setNames(
  as.data.frame(lapply(ordering_names, function(nm) {
    fit_partition$locations[[nm]]$mean$pseudotime
  })),
  ordering_names
)

cor(t_df) |>
  knitr::kable(digits = 3)
```



## Step 4: Refit with known measurement error

The raw file includes paired `*_error` columns, and `fit_mpcurve()` can use
them directly through the `S` argument. In the public API, `S` is interpreted
as a matrix of known measurement standard deviations.

This matters here because ignoring heterogeneous measurement error can make the
model over-explain noisy columns by introducing extra orderings.

### Single-ordering fit with known measurement SD

```{r uni-fit-known-sd}
fit_unidir_known_sd <- fit_mpcurve(
  X,
  S = S,
  method = "isomap",
  K = 50,
  iter = 500,
  rw_q = 2,
  lambda_sd_prior_rate = 1e4,
  verbose = FALSE
)
```

```{r uni-known-sd-convergence, out.width = "100%"}
plot(fit_unidir_known_sd, plot_type = "elbo")
```

```{r uni-known-sd-trajectories, fig.height = 8, out.width = "100%"}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(fit_unidir_known_sd, dims = 1)
plot(fit_unidir_known_sd, dims = 2)
plot(fit_unidir_known_sd, dims = 31)
plot(fit_unidir_known_sd, dims = 45)
par(mfrow = c(1, 1))
```

### Greedy search with known measurement SD

We first try forward selection to see whether the data support multiple orderings once we account for measurement error.
This approach starts with a single ordering and adds orderings until the ELBO stops improving.

```{r partition-fit-known-sd}
fit_partition_known_sd <- fit_mpcurve(
  X,
  S = S,
  method = "isomap",
  intrinsic_dim = 4,
  greedy = "forward",
  ridge = 1e-6,
  K = 50,
  iter = 500,
  rw_q = 2,
  lambda_sd_prior_rate = 1e4,
  verbose = FALSE
)

fit_partition_known_sd
```

It is also useful to check the same question from the other direction by
starting from the upper bound and pruning redundant orderings.

```{r partition-fit-known-sd-backward}
fit_partition_known_sd_backward <- fit_mpcurve(
  X,
  S = S,
  method = "isomap",
  intrinsic_dim = 4,
  greedy = "backward",
  ridge = 1e-6,
  K = 50,
  iter = 500,
  rw_q = 2,
  lambda_sd_prior_rate = 1e4,
  verbose = FALSE
)

fit_partition_known_sd_backward
```

When both forward and backward selection collapse back to a single ordering,
that is strong evidence that the earlier higher-dimensional fit was mainly
absorbing noise heterogeneity rather than revealing genuinely distinct latent
systems.

### Compare the forward and backward single-ordering solutions

```{r compare-known-sd-orderings, out.width = "100%"}
cor(
  fit_partition_known_sd$locations$mean$pseudotime,
  fit_partition_known_sd_backward$locations$mean$pseudotime
)

plot(
  fit_partition_known_sd$locations$mean$pseudotime,
  fit_partition_known_sd_backward$locations$mean$pseudotime,
  xlab = "Latent position (forward search)",
  ylab = "Latent position (backward search)"
)
abline(0, 1, lty = 2)
```

Though both methods find a single ordering.

```{r compare-known-sd-trajectories, fig.height = 8, out.width = "100%"}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(fit_partition_known_sd, dims = 1)
plot(fit_partition_known_sd, dims = 30)
plot(fit_partition_known_sd_backward, dims = 1)
plot(fit_partition_known_sd_backward, dims = 30)
par(mfrow = c(1, 1))
```



## Takeaways

This dataset gives a useful template for real-world `fit_mpcurve()` analyses.

1. Start with a single-ordering fit to establish the basic latent ordering.
2. Use `lambda_sd_prior_rate` when the ordering looks sensible but the fitted
   trajectories are too rough to interpret comfortably.
3. Use `intrinsic_dim` together with greedy search to test whether the data
   support multiple ordering systems.
4. If measurement error is available, pass it through `S`; otherwise the model
   may over-split noisy environments into extra orderings.

