---
title: "Getting Started with MPCurver CAVI"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting Started with MPCurver CAVI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

```{r load}
library(MPCurver)
set.seed(42)
```

`MPCurver` now uses a CAVI backend through the public `fit_mpcurve()` and
`do_mpcurve()` interface. This vignette introduces the statistical model, the
variational algorithm, and then walks through a larger but still easy spiral
example.

## 1  Statistical model

We observe a data matrix
\[
X = (x_{ij}) \in \mathbb{R}^{n \times d},
\]
where row \(i\) is one cell and column \(j\) is one feature.

The single-ordering MPCurve model assumes:

1. each cell has a latent position
   \[
   c_i \in \{1,\dots,K\},
   \]
2. each feature has a smooth trajectory over the \(K\) ordered components,
   written as
   \[
   u_j = (u_{j1},\dots,u_{jK})^\top \in \mathbb{R}^K,
   \]
3. observations follow a Gaussian likelihood
   \[
   x_{ij} \mid c_i = k, u_j, \sigma_j^2 \sim N(u_{jk}, \sigma_j^2),
   \]
4. and each trajectory is regularized by a Gaussian Markov random field prior
   along the component axis:
   \[
   u_j \sim N\!\left(0, (\lambda_j Q_K)^{-1}\right),
   \]
   where \(Q_K\) is the random-walk precision matrix on the \(K\) bins.

In the current `cavi` implementation:

- `sigma_j^2` is feature-specific,
- `lambda_j` is feature-specific,
- the covariance model is homoskedastic across components,
- and the posterior is approximated with a mean-field family
  \[
  q(C,U) = \prod_{i=1}^n q_i(c_i)\prod_{j=1}^d q_j(u_j).
  \]

## 2  Variational algorithm

The backend used by `fit_mpcurve()` performs coordinate-ascent variational
inference on an explicit ELBO. Write
\[
r_{ik} := q_i(c_i = k),
\qquad
N_k := \sum_{i=1}^n r_{ik},
\]
and for each feature \(j\),
\[
q_j(u_j) = N(m_j, S_j).
\]

Under the mean-field family
\[
q(C,U)=\prod_{i=1}^n q_i(c_i)\prod_{j=1}^d q_j(u_j),
\]
the ELBO can be written, up to additive constants, as
\[
\mathcal L(q)
 =
\sum_{i=1}^n\sum_{k=1}^K r_{ik}\log \pi_k
- \sum_{i=1}^n\sum_{k=1}^K r_{ik}\log r_{ik}
\]
\[
- \frac12 \sum_{i=1}^n\sum_{k=1}^K\sum_{j=1}^d
r_{ik}
\left[
\log(2\pi \sigma_j^2)
+
\frac{(x_{ij}-m_{jk})^2 + S_{j,kk}}{\sigma_j^2}
\right]
\]
\[
+ \frac12 \sum_{j=1}^d
\left[
\operatorname{rank}(Q_K)\log \lambda_j
- \lambda_j\Big(m_j^\top Q_K m_j + \operatorname{tr}(Q_K S_j)\Big)
+ \log |S_j|
\right].
\]

This is the quantity recorded in `elbo_trace`. The updates used by the public
backend are obtained by maximizing this ELBO one block at a time.

### 2.1  Update for \(q(u_j)\)

Fix \(r_{ik}\), \(\sigma_j^2\), and \(\lambda_j\). The ELBO terms involving
\(u_j\) are quadratic, so the optimal variational factor is Gaussian:
\[
\log q_j^\star(u_j)
=
-\frac12 u_j^\top A_j u_j + u_j^\top b_j + \text{const},
\]
where
\[
A_j
=
\lambda_j Q_K +
\operatorname{diag}\!\left(N_1/\sigma_j^2,\dots,N_K/\sigma_j^2\right),
\]
and
\[
b_{jk}
=
\frac{1}{\sigma_j^2}\sum_{i=1}^n r_{ik} x_{ij}.
\]
Therefore,
\[
S_j = A_j^{-1},
\qquad
m_j = S_j b_j.
\]

This is exactly the linear-algebra step implemented by the backend behind
`fit_mpcurve()`.

### 2.2  Update for \(q(c_i)\)

Fix \(q(U)\), \(\pi\), and \(\sigma_j^2\). Collecting the ELBO terms involving
the row \(r_{i\cdot}\) gives
\[
\log r_{ik}
\propto
\log \pi_k
- \frac12 \sum_{j=1}^d
\left[
\log(2\pi \sigma_j^2)
+
\frac{(x_{ij}-m_{jk})^2 + S_{j,kk}}{\sigma_j^2}
\right].
\]
So the cell-position posterior is updated by a row-wise softmax:
\[
r_{i\cdot} \leftarrow \operatorname{softmax}\!\big(\ell_{i1},\dots,\ell_{iK}\big),
\]
with \(\ell_{ik}\) equal to the right-hand side above.

The extra \(S_{j,kk}\) term is the uncertainty correction that distinguishes
this update from a pure plug-in EM step.

### 2.3  Update for \(\pi_k\)

Fixing everything else, we maximize
\[
\sum_{k=1}^K N_k \log \pi_k
\]
subject to \(\sum_k \pi_k = 1\). A Lagrange multiplier gives
\[
\pi_k = \frac{N_k}{n}.
\]

### 2.4  Update for \(\sigma_j^2\)

Fixing \(q(U)\) and \(q(C)\), differentiate the ELBO with respect to
\(\sigma_j^2\). The stationary point is
\[
\sigma_j^2
=
\frac{1}{n}
\sum_{i=1}^n \sum_{k=1}^K
r_{ik}\Big[(x_{ij}-m_{jk})^2 + S_{j,kk}\Big].
\]

So each feature variance is updated from the posterior expected residual sum of
squares.

### 2.5  Update for \(\lambda_j\)

Fixing \(q(U)\), only the prior part of the ELBO depends on \(\lambda_j\):
\[
\frac12 \operatorname{rank}(Q_K)\log \lambda_j
- \frac12 \lambda_j
\Big(m_j^\top Q_K m_j + \operatorname{tr}(Q_K S_j)\Big).
\]
Setting the derivative to zero yields
\[
\lambda_j
=
\frac{\operatorname{rank}(Q_K)}
{m_j^\top Q_K m_j + \operatorname{tr}(Q_K S_j)}.
\]

### 2.6  Why the ELBO is monotone here

Each of the five updates above is the exact optimizer of one ELBO block while
holding the others fixed. That is why the current `cavi` implementation has a
clean monotonicity story: the recorded `elbo_trace` should be nondecreasing up
to numerical tolerance.

## 3  Simulated spiral example

We use a larger and more stable spiral than the older vignette examples:

- `n = 1000` cells,
- a shorter latent range `t in [1.5*pi, 4.5*pi]`,
- lower noise,
- and a finer mixture grid `K = 60`.

```{r simulate}
sim <- simulate_swiss_roll_1d_2d(
  n = 1000,
  t_range = c(1.5 * pi, 4.5 * pi),
  sigma = 0.12,
  seed = 1
)

X <- as.matrix(sim$obs)
t_true <- sim$t

cat("Data dimensions:", nrow(X), "x", ncol(X), "\n")
cat("t range:", round(range(t_true), 3), "\n")
```

```{r plot-data, fig.cap="Shorter spiral used in the introductory CAVI example."}
pal <- grDevices::colorRampPalette(
  c("#1F3A5F", "#2E8A99", "#F2C14E", "#E76F51")
)(256)
col_true <- pal[cut(t_true, breaks = 256, labels = FALSE)]

plot(X, col = col_true, pch = 19, cex = 0.55,
     xlab = "x", ylab = "y",
     main = "Observed spiral, coloured by true pseudotime")
```

## 4  Fit the default public interface

```{r fit-cavi}
fit <- fit_mpcurve(
  X,
  method = "fiedler",
  K = 60,
  iter = 120
)

stopifnot(inherits(fit, "mpcurve"))
stopifnot(identical(fit$algorithm, "cavi"))

fit
```

```{r summary-fit}
summary(fit)
```

## 5  Check convergence and recovery

The default `cavi` fit should have a nondecreasing ELBO trace.

```{r check-elbo}
stopifnot(all(diff(fit$elbo_trace) >= -1e-8))
cat("ELBO monotone:", TRUE, "\n")
cat("ELBO range:", round(range(fit$elbo_trace), 3), "\n")
```

We can also compare the inferred ordering to the true latent parameter using
the posterior-mean component index
\[
\hat t_i = \sum_{k=1}^K k \, r_{ik}.
\]

```{r ordering-recovery}
pseudotime_hat <- as.numeric(fit$gamma %*% seq_len(ncol(fit$gamma)))
rho_t <- abs(cor(pseudotime_hat, t_true, method = "spearman"))

cat("Absolute Spearman correlation with true t:", round(rho_t, 4), "\n")
```

## 6  Visualize the fitted ordering

```{r plot-fit, fig.cap="True pseudotime (left) and inferred CAVI pseudotime (right)."}
col_fit <- pal[cut(pseudotime_hat, breaks = 256, labels = FALSE)]

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
plot(X, col = col_true, pch = 19, cex = 0.5,
     xlab = "x", ylab = "y", main = "Truth")
plot(X, col = col_fit, pch = 19, cex = 0.5,
     xlab = "x", ylab = "y", main = "CAVI fit")
par(op)
```

The built-in `mpcurve` plotting methods work directly on the fitted object.

```{r plot-trace, fig.cap="CAVI trace plot from the fitted mpcurve object."}
plot(fit, plot_type = "elbo")
```

```{r plot-mu, fig.cap="Posterior mean trajectories across the K bins."}
plot(fit, plot_type = "mu")
```

## 7  Takeaway

For this easier spiral regime, the public CAVI workflow is:

- fast enough to use as the default frontend algorithm,
- explicit about its optimization target through the ELBO,
- and able to recover the ordering very accurately on a moderately large
  example with a fairly fine grid.
