---
title: "Feature Partitioning with MPCurver"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Feature Partitioning with MPCurver}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

```{r load}
library(MPCurver)
set.seed(42)
```

## The feature-partition problem

The single-ordering model from `vignette("mpcurve_intro")` assumes that all features
share one latent sample ordering. In many applications, however, different
groups of features may reflect **different** underlying processes, each with
its own ordering.

The partition model with $M$ orderings introduces:

- a feature-level assignment $z_j \in \{1,\dots,M\}$,
- $M$ independent latent orderings $t_i^{(m)} \in \{1,\dots,K\}$, $m = 1,\dots,M$,
- and $M$ sets of smooth trajectories $u_j^{(m)} \in \mathbb{R}^K$.

The observation model becomes:
$$
x_{ij} \mid z_j = m,\; t_i^{(m)} = k
\;\sim\;
N\!\bigl(u_{jk}^{(m)},\; \sigma_j^2\bigr).
$$

Each trajectory retains the GMRF prior:
$$
u_j^{(m)} \sim N\!\bigl(0,\; (\lambda_j^{(m)} Q_K)^{-1}\bigr).
$$

The goal is to recover both the **partition** (which features belong to which
ordering) and the **orderings** themselves.

This generalises the `intrinsic_dim` concept from `vignette("mpcurve_intro")`:
a model with $M$ orderings has **intrinsic dimension $M$**.
The standard single-ordering fit is the special case $M = 1$.

## The partition-CAVI algorithm

`MPCurver` uses a soft partition approach: rather than assigning each feature
to exactly one ordering, it maintains probabilistic weights $w_{jm} \in [0,1]$
(with $\sum_m w_{jm} = 1$) describing how much feature $j$ belongs to ordering
$m$.

### Per-feature likelihood scores

For each feature $j$ and ordering $m$, the assignment update uses the
**expected log-likelihood score**
$$
\ell_{jm}^{\mathrm{like}}
=
\mathbb{E}_{q(C^{(m)}) q(u_j^{(m)})}
\big[
\log p(x_{\cdot j} \mid z_j = m,\; C^{(m)},\; u_j^{(m)})
\big],
$$
which expands to
$$
\ell_{jm}^{\mathrm{like}}
=
\sum_{i,k} r_{ik}^{(m)}
  \left[
    -\tfrac{1}{2}\log(2\pi\sigma_j^2)
    - \frac{(x_{ij} - m_{jk}^{(m)})^2 + S_{j,kk}^{(m)}}{2\,\sigma_j^2}
  \right].
$$

### Soft weights via annealed softmax

Under the default fixed-uniform assignment prior, the feature weights are
updated via a softmax at temperature $T$:
$$
w_{jm} \;\propto\; \exp\!\bigl(\ell_{jm}^{\mathrm{like}} / T\bigr).
$$

If `assignment_prior = "dirichlet"`, the score is augmented by the current
variational expectation of the ordering-usage weights:
$$
w_{jm} \;\propto\; \exp\!\left(
\bigl[\ell_{jm}^{\mathrm{like}} + \mathbb{E}_q \log \omega_m\bigr] / T
\right),
$$
with
\[
q(\omega)
=
\operatorname{Dirichlet}\!\left(
\alpha + \sum_j w_{j1},\dots,\alpha + \sum_j w_{jM}
\right).
\]

Each CAVI ordering update uses feature-weighted responsibilities:
the contribution of feature $j$ to ordering $m$ is scaled by $w_{jm}$.

The **annealing schedule** starts at $T = T_\text{start}$ and decreases
geometrically to $T = T_\text{end}$. At high temperature, all features
contribute more evenly across orderings; as $T \to 1$, the weights sharpen
toward hard assignments.

### Full variational objective

Under the default fixed-uniform prior, the partition objective is
$$
\mathcal{L}_\text{partition}
= - d \log M
  + T\,H(w)
  + \sum_{m=1}^M
    \Big[
      \mathbb{E}_q \log p(C^{(m)} \mid \pi^{(m)})
      + H(q(C^{(m)}))
    \Big]
  + \sum_j \sum_{m=1}^M
    \Big[
      w_{jm}\,\ell_{jm}^{\mathrm{like}}
      - \mathrm{KL}\!\bigl(q(u_j^{(m)}) \,\|\, p(u_j^{(m)})\bigr)
    \Big].
$$

With `assignment_prior = "dirichlet"`, the fixed term `-d log M + T H(w)` is
replaced by the full variational assignment block
$$
\sum_{j,m} w_{jm}\,\mathbb{E}_q \log \omega_m
+ T\,H(w)
- \mathrm{KL}\!\bigl(q(\omega) \,\|\, p(\omega)\bigr).
$$

In both cases, if `lambda_sd_prior_rate > 0`, the recorded objective also
includes the induced prior contribution
$$
\sum_{j,m} \log p(\lambda_j^{(m)}),
\qquad
\log p(\lambda) =
\log(\rho/2) - \tfrac{3}{2}\log \lambda - \rho \lambda^{-1/2},
$$
so the stored `objective_history` remains aligned with the actual penalized
updates.

## Fitting via `fit_mpcurve()`

The main entry point for partition models is `fit_mpcurve()` with
`intrinsic_dim = M`. This returns an `mpcurve` object with the full S3
interface, just like a single-ordering fit.

### Simulated example

We simulate a dataset where 8 features follow ordering A and 8 follow
ordering B:

```{r simulate}
sim <- simulate_dual_trajectory(
  n = 500,
  d1 = 10,
  d2 = 10,
  d_noise = 0,
  sigma = 0.05,
  trajectory_family = c("quadratic", "monotone"),
  seed = 1
)

X <- sim$X
truth <- sim$true_assign

cat("Data dimensions:", nrow(X), "x", ncol(X), "\n")
cat("True partition:\n")
print(table(truth))
```

### Fitting with `intrinsic_dim = 2`

```{r fit-soft}
fit_partition <- fit_mpcurve(
  X,
  intrinsic_dim = 2,
  method = "PCA",
  K = 50,
  rw_q = 2,
  verbose = FALSE
)
```

The result is an `mpcurve` object with `intrinsic_dim = 2`, containing:

- `$fits`: list of 2 single-ordering `mpcurve` fits
- `$partition$pi_weights`: $d \times 2$ matrix of soft feature weights
- `$partition$assign`: hard assignment for each feature (`"A"` or `"B"`)
- `$objective_history`: partition objective trace

```{r visualize-mpcurve, fig.cap="Visualising the fitted `mpcurve` object."}
plot(fit_partition, plot_type = "elbo")
plot(fit_partition, dims = 4)
```


### Checking the objective

For a fixed `intrinsic_dim = M`, the recorded partition objective is the main
trace to inspect. Away from freeze events it should behave monotonically under
the phase-2 updates. If an iteration freezes one or more weak orderings, the
objective for that step is still recorded, but that step is skipped for
convergence accounting and can show a small decrease.

```{r check-soft}
has_drop <- any(diff(fit_partition$objective_history) < -1e-8)
cat("Any recorded decreases:", has_drop, "\n")
print(fit_partition$convergence_info)
```

```{r plot-soft-obj, fig.cap="Variational objective over iterations."}
plot(fit_partition$objective_history, type = "b", pch = 19, cex = 0.7,
     xlab = "Iteration",
     ylab = "Variational objective",
     main = "Soft partition objective")
```

The absolute value of `$objective_history` is most useful when comparing fits
under the **same prior settings**. Its interpretation now depends on
`drop_unused_ordering`:

- If `drop_unused_ordering = FALSE`, the stored objective is the
  fixed-requested-`M` comparison objective. Frozen orderings keep their
  preserved feature-assignment weights and cell-ordering block, while their
  trajectory `U` contribution is neutralized to zero. This is the mode to use
  when comparing fits across requested values of `intrinsic_dim`.
- If `drop_unused_ordering = TRUE`, the stored objective is the post-drop
  fitting objective of the active model. It is useful for monitoring that fit,
  but should not be compared across requested values of `intrinsic_dim`.

Even in comparison mode, `$objective_history` is still a variational
objective, not a calibrated marginal likelihood.

### Visualising both orderings

`plot()` shows one panel per ordering, each coloured by that ordering's
pseudotime. The title annotates the mean feature weight for the plotted
dimensions:

```{r plot-partition, fig.width=10, fig.height=5, fig.cap="Both orderings side by side."}
plot(fit_partition, dims = 1)
```

### Partition accuracy

```{r soft-accuracy}
pred_assign <- as.character(fit_partition$partition$assign)
truth_assign <- as.character(truth)
soft_acc <- max(
  mean(pred_assign == truth_assign),
  mean(pred_assign != truth_assign)
)
cat("Soft partition accuracy:", round(soft_acc, 4), "\n")
```

Compute the correlation between the inferred pseudotimes and the true orderings:

```{r soft-correlation}
cor1 <- cor(fit_partition$fits[[1]]$locations$mean$pseudotime , sim$t2)
cor2 <- cor(fit_partition$fits[[2]]$locations$mean$pseudotime, sim$t1)
cat("Correlation with true orderings:\n")
cat("Ordering 1:", round(cor1, 4), "\n")
cat("Ordering 2:", round(cor2, 4), "\n")
```


### Continuing a fit

Use `do_mpcurve()` to run additional convergence iterations at $T = T_\text{end}$:

```{r continue, eval=FALSE}
fit_partition <- do_mpcurve(fit_partition, iter = 10)
```

This extends `$objective_history` in place, just like for single-ordering fits.

## Initialisation and methods

By default, partition fits now use `partition_init = "similarity"`, which first
clusters features and then initializes one ordering per feature block.

If you explicitly choose `partition_init = "ordering_methods"`, ordering 1 uses
the `method` argument (same as `fit_mpcurve()` for `intrinsic_dim = 1`), and
orderings 2, 3, … use successive PCA components (PC2, PC3, …) to ensure
diverse initialisation rather than duplicating the first axis.

You can specify independent methods for each ordering:

```{r fit-methods, eval=FALSE}
# Ordering 1: fiedler, ordering 2: PCA (PC1)
fit2 <- fit_mpcurve(X, intrinsic_dim = 2,
                    method = c("fiedler", "PCA"),
                    K = 16)
```

For $M > 2$, extend the vector accordingly:

```{r fit-m3, eval=FALSE}
# Three orderings with different initialisations
fit3 <- fit_mpcurve(X, intrinsic_dim = 3,
                    method = c("fiedler", "PCA", "PCA"),
                    K = 16)
```

## Greedy dimension selection

Dimension selection is now unified under the partition-CAVI workflow through
`fit_mpcurve(greedy = ...)`.

- `greedy = "forward"` treats `intrinsic_dim` as an upper bound and compares
  `M` versus `M + 1` sequentially starting from `M = 1`.
- `greedy = "backward"` first fits the upper bound, then removes one active
  ordering at a time and compares `M` versus `M - 1`.

Both greedy modes always compare models using the fixed-requested-`M`
comparison objective, so they internally run with `drop_unused_ordering = FALSE`
even if you ask for a compact final returned view.

```{r fit-greedy}
fit_forward <- fit_mpcurve(
  X,
  intrinsic_dim = 4,
  greedy = "forward",
  K = 16,
  drop_unused_ordering = FALSE,
  verbose = FALSE
)

fit_backward <- fit_mpcurve(
  X,
  intrinsic_dim = 4,
  greedy = "backward",
  K = 16,
  drop_unused_ordering = TRUE,
  verbose = FALSE
)
```

```{r summary-table}
summary_tbl <- data.frame(
  fit = c("soft (fixed M=2)", "greedy forward", "greedy backward"),
  displayed_dim = c(
    fit_partition$intrinsic_dim,
    fit_forward$intrinsic_dim,
    fit_backward$intrinsic_dim
  ),
  selected_dim = c(
    fit_partition$requested_intrinsic_dim,
    fit_forward$greedy_selection$selected_intrinsic_dim,
    fit_backward$greedy_selection$selected_intrinsic_dim
  )
)
print(summary_tbl, row.names = FALSE)
```

```{r helpers, echo=FALSE}
assignment_heatmap <- function(truth, ..., main = "Feature partitions") {
  mats <- list(...)
  ord <- order(truth)
  encode <- function(x) ifelse(x[ord] == "A", 1, 2)
  Z <- do.call(rbind, c(list(truth = encode(truth)), lapply(mats, encode)))

  cols <- c("#4C78A8", "#F58518")
  op <- par(mar = c(4, 6, 2, 1))
  on.exit(par(op), add = TRUE)
  image(
    x = seq_len(ncol(Z)),
    y = seq_len(nrow(Z)),
    z = t(Z[nrow(Z):1, , drop = FALSE]),
    col = cols,
    axes = FALSE,
    xlab = "Features (sorted by truth)",
    ylab = "",
    main = main
  )
  axis(2, at = seq_len(nrow(Z)), labels = rev(rownames(Z)), las = 1)
  box()
}
```

```{r plot-assignments, fig.cap="True vs inferred feature partitions."}
assignment_heatmap(
  truth    = truth,
  soft     = fit_partition$partition$assign,
  forward  = fit_forward$partition$assign,
  backward = fit_backward$partition$assign,
  main = "Feature partition comparison"
)
```

## Key parameters

### `fit_mpcurve()` with `intrinsic_dim >= 2`

| Parameter | Default | Description |
|---|---|---|
| `intrinsic_dim` | `1` | Number of orderings $M$; in greedy modes this is an upper bound |
| `greedy` | `"none"` | Optional dimension-selection mode: `"forward"` or `"backward"` |
| `method` | `"PCA"` | Length-1 or length-$M$ vector of initialisation methods per ordering |
| `K` | auto | Number of grid points per ordering |
| `n_outer` | `25` | Annealing iterations ($T$ decreases geometrically) |
| `max_converge_iter` | `100` | Post-annealing iterations at $T = T_\text{end}$ |
| `tol_outer` | `1e-5` | Phase-2 relative objective tolerance; requires 3 consecutive small nonnegative steps |
| `T_start` / `T_end` | `5` / `1` | Temperature schedule endpoints |
| `inner_iter` | `1` | Weighted CAVI sweeps per outer step |

## Tips

- **Use `fit_mpcurve(intrinsic_dim = M)` as the main entry point.** It returns
  a unified `mpcurve` object and handles all initialisation automatically.
- **Use `greedy = "forward"` or `"backward"` when `intrinsic_dim` is only an upper bound.**
  Greedy search always compares models with the fixed-`M` objective.
- **Start with `intrinsic_dim = 2`.** If the partition is unclear, consider
  $M = 3$ to allow a third ordering to capture residual structure.
- **Specify independent init methods.** `method = c("fiedler", "PCA")` uses a
  spectral ordering for the first axis and PC1 for the second explicit
  PCA-based ordering. If you want later PCA axes as explicit starts, pass
  `pca_components` directly.
- **Non-monotone features are harder.** When many features have non-monotone
  relationships with the latent ordering, results become more
  initialisation-sensitive.
- **Partitioning is easier than trajectory recovery.** Even when the inferred
  orderings are imperfect, the feature partition can still be accurate.
- **Check the objective trace.** Within one fixed `intrinsic_dim`, sustained
  decreases are a red flag; a one-step drop on a freeze-affected iteration is
  possible and is excluded from convergence accounting.
- **Compare across `intrinsic_dim` only with `drop_unused_ordering = FALSE`.**
  In that mode the stored objective keeps the fixed requested-`M` comparison
  semantics. With `drop_unused_ordering = TRUE`, it becomes a post-drop fit
  objective and should not be compared across requested dimensions.

## Summary

- `fit_mpcurve(X, intrinsic_dim = M)` is the primary entry point for partition
  models, returning an `mpcurve` with `intrinsic_dim = M`.
- `fit_mpcurve(..., greedy = "forward")` and
  `fit_mpcurve(..., greedy = "backward")` provide the unified dimension
  selection workflow going forward.
- `print()`, `summary()`, and `plot()` all work out of the box:
  `plot()` shows $M$ panels, one per ordering, coloured by pseudotime.
- `do_mpcurve()` continues fitting at $T = T_\text{end}$ and extends the
  objective history, identical to the single-ordering interface.
- All methods build on the same CAVI trajectory model from `vignette("mpcurve_intro")`.
